An AI is haunting campus

### 2024-06-19 ai-haunting-campus/ai-haunting-campus.qmd
First full draft completed

@@

As I argue above, AIs do not seem to do much for students or instructors, other than distract them from learning and developing genuinely transferable skills. On the other hand, it is hard not to suspect that the provincial government is hoping to use GenAIs to further reduce the need to fund higher education. But AIs must do something for someone, otherwise why would they be everywhere? For Ontario, the main patron and effectively the boss of McMaster, it provides a rationale (an excuse even) to force the institution to find "efficiencies". Tom Goldstein, a computer science professor at the University of Maryland, thinks that the lure of GenAI will be hard to resist for many, irrespective of their flaws...after all, "[e]ven though [GenAI is] expensive...it's still far less expensive than human labor" [@oremus_chatgpt_2023]. Without adjudicating the merits of the argument that human labor is expensive^[When the alternative is substandard quality of life for many of our fellow citizens, and/or higher unemployment.], it is hard to argue against the lure of GenAI for governments that do not seem to believe in higher education.

@@

@perrigo_2_2023 (Follows summary)

ChatGPT was trained on vast amounts of data scooped from the internet, where many societal problems are amplified. Trained on horrible content, it was hard to sell ChatGPT when it was prone to racist, sexist, obscene outbursts. But companies like facebook, who did not want to pay for moderation, had hit on a solution of sorts <!--: train an AI to detect content that was unpalatable.-->

The solution?

> [F]eed an AI with labeled examples of violence, hate speech, and sexual abuse, and that tool could learn to detect those forms of toxicity in the wild. That detector would be built into ChatGPT to check whether it was echoing the toxicity of its training data, and filter it out before it ever reached the user. It could also help scrub toxic text from the training datasets of future AI models.

The key here is "labelled examples". What are those labels? Many machine algorithms do not learn by themselves...they learn from humans, and need what is called "labelled" data: a images that humans have labelled as "dog", "cats", etc. Or bits of text that have been labelled as...well...questionable. Labeling data is extremely time consuming and labor intensive. What's a corporation to do?

> To get those labels, OpenAI sent tens of thousands of snippets of text to an outsourcing firm in Kenya, beginning in November 2021. Much of that text appeared to have been pulled from the darkest recesses of the internet. Some of it described situations in graphic detail like child sexual abuse, bestiality, murder, suicide, torture, self harm, and incest. "

What did this mean to the people doing all this labeling in one of the poorest regions in the world? First of all, an income: "data labelers employed...on behalf of OpenAI were paid a take-home wage of between around $1.32 and $2 per hour depending on seniority and performance." But at what cost?

> One Sama worker tasked with reading and labeling text for OpenAI told TIME he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child. “That was torture,” he said. “You will read a number of statements like that all through the week. By the time it gets to Friday, you are disturbed from thinking through that picture."

@@

> Karen Hao: You can interact with ChatGPT in Swahili, but ChatGPT was developed primarily to work in English. So a lot of the scrubbing, the content moderation, the important safety measures within the chatbot were done in English. So when you prompt it in Swahili, you'll get more misinformation. You'll get more confusing sentences that don't make sense, and you will potentially get more of this content that they worked so hard to filter out because they were only filtering it in English.

@roberts_commercial_2016 Commercial {Content} {Moderation}: {Digital} {Laborers}' {Dirty} {Work}

### Copyright (and privacy) issues

@@

Neural networks and machine learning are nothing new, and they have their uses. But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI. This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.]. There are now calls to press pause on the hype [@angwin_press_2024] but they are barely audible above the reverberating echos inside the AI hype chamber.

@@

In this essay I have tried to describe the many problems with AI as an _everything tool_. Tall in my mind are the moral and ethical concerns around an extractive and exploitative technology that promises democracy but delivers further concentration of wealth and power.

<!--
> An end to “batch” teaching. Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning. Where does the student want the learning to take them? AI systems, coupled with coaching, mentoring and some instruction and peer support networks then ensure the learner is supported on the carefully mapped learning journey. Teaching can still take place: Students challenged with a set of related issues can be brought together for a boot camp or focused period of instruction before returning to their individualized study activities.

@@

> Greater accessibility. Higher education systems are organized around blocks of time: semesters or terms. There is no longer good reason for this. Students should be able to begin their learning at any time and be assessed when they are ready. A four-year undergraduate degree could easily be completed in 16 months by a diligent student, for example, or in even less time by someone arriving in the program with substantial life and work experience. AI systems can also leverage technologies to ensure the resources and supports for learners reflect who they are (culture, language, experience and skill) and what challenges they face (disabilities, learning styles, etc.).
-->

It has now been 55 years since Howard S. Barrows and Robyn M. Tamblyn [pioneered](https://en.wikipedia.org/wiki/Problem-based_learning#History) problem-based learning at McMaster, an accomplishment that the university [still touts today](https://provost.mcmaster.ca/teaching-learning/). Problem-based learning was not and is still not high tech, and yet remains a great pedagogical approach, influential to the date. Developing problem-based learning took human ingenuity and insight, not getting hooked on an expensive technology of questionable value today, of dubious longevity, and that in the here and now is unarguably extractive, abusive, and exploitative, and that advances an agenda that can only be squared with "integrity, quality, and excellence" via fabulous contortions of the imagination.

Does the university imagine that it will be best known in half a century as "McMaster, the institution that pioneered the use of ChatGPT"? Is that what it really wants? [Regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean), after all, is also called a "reversion to mediocrity" [@barnett_regression_2017], and that is what an AI campus has to offer.

I suggest that universities should avoid the mistakes of the media, an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) [also see the "pivot-to-video" and how tech companies suckered many media companies into giving them money in exchange for fake engagement statistics [@hazard_owen_facebooks_2021].]. Tech corporations (or least ways some of them), can still hope to recoup some of their investment in AI. But a university gambles its most important asset when betting on AI: its reputation as a place were knowledge is created, and as a custodian of knowledge. AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Entities like Google and Amazon could not care less about their reputation. They are [too big](https://www.businessinsider.com/ex-amazon-ghaderi-exec-suing-ai-race-copyright-allegations-2024) [to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/).] to the unaware^[Like a B.C. lawyer who was [reprimanded](https://www.cbc.ca/news/canada/british-columbia/lawyer-chatgpt-fake-precedent-1.7126393) for using AI hallucinations in a legal filing. This was not even the first time that supposedly highly skilled professionals were castigated for [using](https://www.washingtonpost.com/technology/2023/11/16/chatgpt-lawyer-fired-ai/) [chatbots](https://www.reuters.com/legal/new-york-lawyers-sanctioned-using-fake-chatgpt-cases-legal-brief-2023-06-22/).], and beyond, to bad-faith actors^[Like rival geopolitical powers that are relatively impervious to propaganda themselves for reason of being [authoritarian countries](https://www.theatlantic.com/magazine/archive/2024/06/china-russia-republican-party-relations/678271/).]. In my view, there are no efficiencies, no promises of future excellence, that can compensate the risk of ruining an institutions reputation.

Why? The reputational risk...universities are under siege from unsympathetic or actively hostile actors. Universities main source of power is their reputation as places that preserve and expand knowledge in a principled way. "This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an institution of higher-education is not needed when there are cheaper alternatives for such an education ("act in the next four days to proof yourself against layoffs!"):
<!--
![](images/trail-blazing-ai-bundle.png)
-->

@@

AI industry and the military: @haskins_im_2024

## Bonus

```{r googly-eyes, out.width="70%", fig.cap="\\label{fig:googly-ai-eyes}In conclusion"}
google_plot <- ggplot() +
  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
  theme_void()  +
  theme(text = element_text(size = 8))

>>>>

License: [CC-BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/)

![](by-sa.png){width=50}

@@

As I argue above, AIs do not seem to do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.

But AIs must do something for someone, otherwise why would they be everywhere? Given history, one could be forgiven to think that the provincial government is hoping to use GenAIs to further reduce the need to fund higher education. GenAIs appear on the horizon as an  innovation-sounding rationale (a mandate even!) for the province to force institutions of higher education to find "efficiencies". Tom Goldstein, a computer science professor at the University of Maryland, thinks that the lure of GenAI will be hard to resist for many, irrespective of the technology's flaws...after all, "[e]ven though [GenAI is] expensive...it's still far less expensive than human labor" [@oremus_chatgpt_2023]. Without adjudicating the merits of the argument that human labor is expensive^[When the alternative is substandard quality of life for many of our fellow citizens, and/or higher unemployment.], one would hav to be blind not to see the glamour that GenAI will cast on governments that do not seem to believe in higher education in the first place.

```{r googly-eyes, out.width="70%", fig.cap="\\label{fig:googly-ai-eyes}'I like the look of his cuts'"}
google_plot <- ggplot() +
  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
  theme_void()  +
  theme(text = element_text(size = 8))

@@

@@

> Karen Hao: You can interact with ChatGPT in Swahili, but ChatGPT was developed primarily to work in English. So a lot of the scrubbing, the content moderation, the important safety measures within the chatbot were done in English. So when you prompt it in Swahili, you'll get more misinformation. You'll get more confusing sentences that don't make sense, and you will potentially get more of this content that they worked so hard to filter out because they were only filtering it in English.
-->

### Privacy issues

@@

Neural networks and machine learning are nothing new, and they have their uses. But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI. This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.]. There are now calls to press pause on the hype [@angwin_press_2024] but these voices are barely audible above the thundering echos inside the AI hype chamber.

@@

In this essay I have tried to describe the many problems with AI as an _everything tool_. Tall in my mind are the moral and ethical concerns around an extractive and exploitative technology that promises democracy but delivers further concentration of wealth and power. GenAI is environmentally damaging for what it delivers: confidently stated confabulations. I have little illusion that these arguments will prove persuasive, since the university is already aware of them because they usually appear in the last slides of every presentation about--and yet here we are, asking not whether, but where it makes sense to adopt AI.

I would still appeal, though, to the university's self-regard and sense of self-preservation.

It has now been 55 years since Howard S. Barrows and Robyn M. Tamblyn pioneered [problem-based learning](https://en.wikipedia.org/wiki/Problem-based_learning#History) at McMaster, an accomplishment that the university [still touts today](https://provost.mcmaster.ca/teaching-learning/). Problem-based learning was not and is still not high tech, and yet remains a remarkable pedagogical breakthrough, an approach that has stayed influential for decades. Developing problem-based learning took human ingenuity and insight. What the university is considering today is which of our core missions can we hitch to an expensive technology of questionable value today, of dubious longevity, and that in the here and now advances an agenda in education that can only be squared with "integrity, quality, and excellence" via fabulous contortions of the imagination.

Does the university imagine that it will be best known in half a century as "McMaster, the institution that pioneered the use of ChatGPT"? Is that what it really wants? All that AIs offer, after all, is a very sophisticated [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean), which not by coincidence is also called a "reversion to mediocrity" [@barnett_regression_2017]. We would do our students an unforgivable disservice by offering a skill that reversed them to mediocrity.  "This university uses ChatGPT to teach" is simply a way of saying that we are too cheap to offer a human-centered education. But in truth, a university is too expensive an institution to be attractive under such conditions, when there are far cheaper alternatives for an AI-based education ("for only $9.99 proof yourself against layoffs").

```{r ai-bundle, out.width="110%", fig.cap="\\label{fig:trail-blazing-bundle}A trail-blazing bundle: ChatGPT for data science and data analysis in Python"}
ggdraw() + 
  draw_image(image = glue::glue(here::here(), "/images/trail-blazing-ai-bundle.png", width = 0.8))
```

Going down this path would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors, a university's main source of authority is its reputation as a place that preserves and expands knowledge in a principled way.

I suggest that universities would do well to avoid the mistakes of other sectors, including those in the media that have contributed to hype AI. The media is an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) [also see the "pivot-to-video" and how tech companies suckered many media companies into giving them money in exchange for fake engagement statistics [@hazard_owen_facebooks_2021].]. Tech corporations (or at least some of them), can still hope to recoup some of the fabulous ammounts they have poured into AI. But a university gambles its most important asset when betting on AI: its reputation as a place were knowledge is created, and as a custodian of knowledge. AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Entities like Google and Amazon could not care less about their reputation. They are [too big](https://www.businessinsider.com/ex-amazon-ghaderi-exec-suing-ai-race-copyright-allegations-2024) [to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/).] to the unaware^[Like a B.C. lawyer who was [reprimanded](https://www.cbc.ca/news/canada/british-columbia/lawyer-chatgpt-fake-precedent-1.7126393) for using AI hallucinations in a legal filing. This was not even the first time that supposedly highly skilled professionals were castigated for [using](https://www.washingtonpost.com/technology/2023/11/16/chatgpt-lawyer-fired-ai/) [chatbots](https://www.reuters.com/legal/new-york-lawyers-sanctioned-using-fake-chatgpt-cases-legal-brief-2023-06-22/).], and beyond, to bad-faith actors^[Like rival geopolitical powers that are relatively impervious to propaganda themselves for reason of being [authoritarian countries](https://www.theatlantic.com/magazine/archive/2024/06/china-russia-republican-party-relations/678271/).].

In my view, there are no efficiencies, no questionable promises of future excellence, that can compensate the one thing that makes a university stand tall among other institutions--its pursuit of innovation, yes, but above all of the truth, because what else is worth discovering, preserving, and communicating?

<!--
> Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning. Where does the student want the learning to take them? AI systems, coupled with coaching, mentoring and some instruction and peer support networks then ensure the learner is supported on the carefully mapped learning journey. Teaching can still take place: Students challenged with a set of related issues can be brought together for a boot camp or focused period of instruction before returning to their individualized study activities.

@@

> Greater accessibility. Higher education systems are organized around blocks of time: semesters or terms. There is no longer good reason for this. Students should be able to begin their learning at any time and be assessed when they are ready. A four-year undergraduate degree could easily be completed in 16 months by a diligent student, for example, or in even less time by someone arriving in the program with substantial life and work experience. AI systems can also leverage technologies to ensure the resources and supports for learners reflect who they are (culture, language, experience and skill) and what challenges they face (disabilities, learning styles, etc.).

@@

AI industry and the military: @haskins_im_2024
-->
